An Extensible Approach to Verification of Embedded Network Systems*

### 2014-04-25 tech-project/implementation.tex
adding led demo tex

\subsection{C Translation Model}
One of the primary challenges in translating from RESOLVE to C is finding a suitable C analog for each RESOLVE module and the constructs allowable in each. Indeed, since we are dealing with an environment where functional correctness is a primary concern, it is important that the code generated by our tool represents as closely as possible the original RESOLVE source. In an effort to make such considerations, at the highest level, the C code we generate makes special considerations for facilities, concepts, and realizations. This scheme is depicted in Figure \ref{fig:relationship} and briefly summarized in Sections \ref{sec:conceptoverview} - \ref{sec:facilitiesrealizations}.

\begin{figure}
\begin{center}
\includegraphics[scale=.60]{figs/relationship.pdf}
\end{center}
\caption{Relationship between RESOLVE module types and the C code generated from each.}
\label{fig:relationship}
\end{figure}

\subsubsection{Concepts}
\label{sec:conceptoverview}
Concept modules produce a single .h file, which provides function pointers for the operations specified in the original concept, as well as structs representing each user defined type.

\subsubsection{Facilities}
\label{sec:facilitiesoverview}
Facilities produce an .h/.c pair: The header .h declares methods that are intended to manage the creation and destruction of all global variables used within the facility, while the .c provides an implementation of these methods.  Note that these create and destroy methods are only responsible for freeing \textit{global} variables. All other translated local functions are responsible for deallocating their own local variables.

\subsubsection{Realizations}
\label{sec:facilitiesrealizations}
We treat realizations of concepts and enhancements slightly different than facility modules. While a .h/.c pair is still produced, the create method designated in the header for realizations is designed to create instances of all types specified by the concept, while the destroy method deallocates these types -- as opposed to simply destroying user created globals.

@@

\begin{figure*}
\begin{center}
\includegraphics[scale=.55]{figs/ast_traversal2.pdf}
\end{center}
\caption{The general flow of information from C translators walk of the AST (first), to user defined templates (middle), ending with formed output (last).}
\label{fig:ast}
\end{figure*}

@@

The C code generation process takes place over the course of a traversal of RESOLVE's abstract syntax tree (AST). The traversal mechanism used is a derivative of the visitor pattern that provides a pre post traversal over all nodes in the tree.

@@

To illustrate the general process of producing runnable C from RESOLVE code, consider the following dummy operation:

\begin{verbatim}
Operation Nothing(); Procedure
        Var X : Integer;
        X := 3;
end Nothing;
\end{verbatim}

%%LOOK AT ME
%This separation between the RESOLVE source tree and the view, allows us to easily modify and tweak the code produced by our tool and allows us easily represent arbitrarily complicated output

Shown in Figure \ref{fig:ast} is a high level depiction of steps taken in translating this operation to C. The first box depicts the AST of \texttt{Nothing}, where nodes are represented with boxes labeled by the constructs they contain. Throughout the walk of this tree, useful information (such as the operation's name, ``Nothing") are extracted from the nodes containing these constructs, and added to the externally defined string-template\footnote{A template can simply be thought of as a ``document with holes" which the user choses when and how to fill}, \texttt{c\_function\_def} template.

In the context of RESOLVE to C translation, these templates, when filled during the aforementioned pre-post visitor traversal of RESOLVE's AST, help simplify the task of producing complicated, arbitrarily nested blocks of structured C output by keeping translation logic strictly within the C translator, and output logic strictly within the templates.

That is, the only actual work being performed within the C translator is forwarding information gathered from individual treenodes, to a series of externally defined templates. This allows us to exploit (in design pattern parlance) a strict model view controller (MVC) separation in the translator's codebase between the mechanism that does the AST visiting (controller), the tree nodes from which we're adding information to templates (model), and the external file containing all available C language templates which shape our output (view).

We feel this approach lends itself the challenge discussed in this paper, as this separation allows us to easily iterate changes to our generated C code without needing to concern ourselves with a line of Java inside the compiler itself. This allows us to easily tweak code and make arbitrarily complicated changes and optimizations to the code -- some of which are detailed later in Section \#.

% general flow Each of these calls are received in the translator in the order in which they are visited within the tree. It is up to the client (in this case, the author of the C-translator) to decide which of these methods they wish to override and perform custom actions within.

@@

\subsection{Memory Allocation}

Dynamic memory allocation is not typically the first choice for most embedded applications. With the telos mote itself  limited to 128 bytes of RAM, hardware and memory constraints such as these, developers of embedded applications tend to favor static memory allocation over dynamic. Many programs, however benefit greatly from dynamic memory allocation not only in terms of clarity and straightforward to be used, including RESOLVE, in order to make it extensible. Previous RESOVE translations to C required static memory allocation \cite{regula:2010}. In this section a stack based dynamic memory allocator is introduced.

@@

The \texttt{salloc()} is a first fit memory allocator. Rather than allocating on the heap, \texttt{salloc()} uses the stack. At compilation, the allocator provisions a fixed size of memory. It requires a small section of meta-data called a block which contains information of about the size of memory allocated, neighboring blocks, as well if the block is free or not. A sample representation of the stack shown in figure \ref{fig:stack}, is a typical example of allocated memory on the stack.

\begin{figure}[!htb]
\centering
\includegraphics[scale=.55]{figs/stack.pdf}
\caption{Representation of memory stack}
\end{figure}
\label{fig:stack}

@@

A memory allocator must provide a mechanism to release, or free memory in order to indicate that it is not being used and can be reallocated for something else. The \texttt{sfree()} operation is an analogous implementation to the standard C \texttt{free()} function for the stack.

\begin{figure}[!htb]
\centering
\includegraphics[scale=.55]{figs/sfree.pdf}
\caption{sfree to free memory}
\end{figure}
\label{fig:free}

\subsubsection{Optimizations}

A common problem that can occur in memory allocation is fragmentation.  As shown in figure \ref{fig:fragmentation}, fragmentation can lead to inefficient and increased memory usage. This problem is magnified on embedded systems with limited memory capabilities. Simple optimizations can be made however to reduce fragmentation. Splitting is one optimization that \texttt{salloc()} to maximize memory usage. Shown in figure \ref{fig:split}, blocks can be split to the size that is required. Another means of optimizing memory usage is joining together freed blocks. When a call to \texttt{sfree()} is made, neighboring blocks are coalesced together, as shown in figure \ref{fig:fuse}.

\begin{figure}[!htb]
\centering
\includegraphics[scale=.55]{figs/fragmentation.pdf}
\caption{fragmentation}
\end{figure}
\label{fig:fragmentation}

\begin{figure}[!htb]
\centering
\includegraphics[scale=.55]{figs/split.pdf}
\caption{block splitting}
\end{figure}
\label{fig:split}

\begin{figure}[!htb]
\centering
\includegraphics[scale=.55]{figs/fuse.pdf}
\caption{block fusing}
\end{figure}
\label{fig:fuse}
%A stack-based approached to dynamic memory allocation requires a fixed size of memory to be set prior to compilation. 
%To date there have been very few practical applications of verifiable code on physical media.

\subsection{An LED Implementation}

\begin{verbatim}
void LED_Telos_create() {
		Std_Boolean_Fac_create();
		Std_Integer_Fac_create();
		Std_Clock_Fac_create();
		r_type_ptr __arg_0 = Std_Integer_Fac_Var.core->createFromInteger(4, Std_Integer_Fac_Var.core->Integer);
		LED_Telos_Facility_Var.core = new_Std_LED_Realiz_for_Leds_Template(__arg_0);
		LED_Telos_Facility_Var.Toggling_Capability = new_Toggling_Capability_of_Leds_Template(LED_Telos_Facility_Var.core);
		Std_Integer_Fac_Var.core->Integer->destroy(__arg_0, Std_Integer_Fac_Var.core->Integer);
}
\end{verbatim}

>>>>

%In this section we discuss each of these phases using the LED component developed in Section \ref{sec:specifiying}.

\subsection{C Translation Model}
One of the primary challenges in translating from RESOLVE to C is finding a suitable C analog for each RESOLVE module and the constructs allowable in each. Indeed, since we are dealing with an environment where functional correctness is a primary concern, it is important that the code generated by our tool represents as closely as possible the original RESOLVE source. In an effort to make such considerations, at the highest level, the C code we generate makes special considerations for facilities, concepts, and realizations.

\begin{figure}
\begin{center}
\includegraphics[scale=.55]{figs/relationship.pdf}
\end{center}
\caption{Relationship between RESOLVE module types and the C code generated from each.}
\label{fig:relationship}
\end{figure}

\subsubsection{Concepts}
Concepts produce a single .h that provide function pointers for each operation specified by the original concept, as well as structs for each user defined type.

\subsubsection{Facilities}
Facilities produce an .h/.c pair: The header .h declares methods that are intended to manage the creation and destruction of all global variables used within the facility, while the .c provides an implementation of these methods.  Note that the create and destroy methods are only responsible for freeing \textit{global} variables, all other translated, local functions are responsible for deallocating their own variables.

\subsubsection{Realizations}
We treat realizations of concepts and enhancements slightly different than facility modules. While a .h/.c pair is still produced, the create method for realizations is designed to create instances of all types specified by the concept, while the destroy method deallocates these types.

%The .h/.c pair generated for realizations include

%While facilities, concept realizations, and enhancement realizations in general produce an .h/.c pair, the .h produced for facility modules differs slightly from the others: Whereas realizations of concepts and enhancements provide

% concept realizations provides methods for creating and destroying a single instance of the type declared by the concept it realizes. However, header files for realizations of facility modules provide methods creating and destroying all global variables within the facility. Common global variables include the set of standard facilities provided to every module in addition to any locally defined within the scope of the facility.

% responsible for declaring functions that create and destroy the objects used

@@

@@

Translation itself is performed over the course of a traversal of RESOLVE's abstract syntax tree (AST). The traversal mechanism used is a derivative of the visitor patter that provides a pre post traversal over all nodes in the tree.

@@

%To make this more concrete, consider the following dummy operation.

%\begin{verbatim}
%Operation Nothing(); Procedure
%        Var X : Integer;
%        X := 3;
%end Nothing;
%\end{verbatim}

%Shown in Figure \ref{fig:ast} is the AST representation of operation \texttt{Nothing}. Here, language constructs are represented as labeled boxes, while the actual traversal over these constructs -- and the order in which it is performed -- is communicated on the right via the call stack. Each of these calls are received in the translator in the order in which they are visited within the tree. It is up to the client (in this case, the author of the C-translator) to decide which of these methods they wish to override and perform custom actions within.

%We feel this particular traversal pattern lends itself to task of source to source translation for the following reasons:

%\begin{itemize}
%\item A visit method for a construct provides a convenient encapsulation of all the logic required to translate RESOLVE construct $x$ to appropriate C construct $y$.

%\item Overriden visit methods apply to every instance of a construct -- meaning RESOLVE's C translator requires few (if any) loops, as the walk itself serves as the method of iteration.

%\item Finally, only the visit methods for the constructs we currently wish to process need to be overridden. This makes it easy to tweak and optimize the size of the translator's codebase, implementing only visit method for language constructs that explicitly require translation.  
%\end{itemize}

@@

%One of the primary allures of this approach to translation is that that adheres to strict MVC design principles -- meaning that all output logic is distinctly separated from translation related logic.

%We take advantage of these pre post methods using templates and simple stack to implement

%Going even further, we demonstrate this by showing the steps taken to translate a simple operation declaration \texttt{nothing} to C.

% for each construct to remain encapsulated within that method, and (almost) completely eliminates the need for complex loop or iteration mechanisms.

%This particular traversal pattern lends itself well to the language translation challenge (specifically source to source) where the pre methods align naturally with beginning of a construct, and conclude with its corresponding post method.  For instance, in the case above, this allows us, upon receiving a callback for In an effort to keep logic implementing our translator distinct and separate from the output language, we used a model view controller framework for implementing

%To illustrate this, consider once again facility \texttt{DoNothing}.

%%%%%%%%%%%%%%%%%%%%%%%
%           Memory Allocation Section              %
%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Memory Allocation}\label{sec:mem}

Memory limitations on embedded hardware has made dynamic memory allocation a difficult, or impossible practice. Many existing programs, including RESOLVE, however,  inherently use heap-based memory. Previous attempts at a RESOLVE to C translation required a constant size for each variable declaration\cite{regula:2010}. This approach can provide efficient use of memory for embedded systems, it does not, however, provide a straight forward approach to translation. In this section, we introduce a RESOLVE wrapped dynamic memory allocator implemented on the stack.

@@

The function \texttt{salloc()}, uses a first fit approach for allocation on the stack, rather than conventional, heap-based allocators. This approach requires that a fixed size of memory is chosen at compilation. In addition, \texttt{salloc()}, requires a section of meta-data, which is denoted as a \texttt{block}, for each record in the memory pool. A \texttt{block} holds referential information of neighboring blocks, the size of the record the block maintains, and if the block is free or is in use. Figure~\ref{fig:stack} is an example of the stack after several calls to \texttt{salloc()}.

\begin{figure}[!htb]
%\centering
\includegraphics[scale=.55]{figs/stack.pdf}
\caption{Representation of memory using salloc()}
Memory allocated with salloc() allocates a block in front of data allocated. Blocks point to their immediate neighbors and hold their size and usage.
\label{fig:stack}
\end{figure}

@@

A memory allocator must provide a mechanism to release, or free memory in order to indicate that it is not being used and can be reallocated. The \texttt{sfree()} function provides deallocation for memory allocated with \texttt{salloc()}. Figure~\ref{fig:free} shows an example set of memory before and after calling \texttt{sfree()}.

\begin{figure}[!htb]
%\centering
\includegraphics[scale=.55]{figs/sfree.pdf}
\caption{sfree to free memory}
\label{fig:free}
\end{figure}

\subsubsection{Optimizing Memory Usage}

A common problem that can occur in memory allocation is fragmentation. During execution, a program can allocate and deallocate an arbitrary number of times. Figure~\ref{fig:fragmentation} shows problems that result from this. This problem is magnified on embedded systems due to their limited memory capacities. There are simple optimizations that can reduce fragmentation.

Block splitting, as shown in Figure~\ref{fig:split}, occurs during allocation. It allows blocks of greater size to be partitioned into the size requested by the allocator.

Fusing blocks is another technique used when \texttt{sfree()} is called. When memory is deallocated, neighboring free blocks are coalesced to form a single large block, as shown in Figure~\ref{fig:fuse}.

\begin{figure}[!htb]
%\centering
\includegraphics[scale=.55]{figs/fragmentation.pdf}
\caption{fragmentation}
\label{fig:fragmentation}
Performing many naive allocation and deallocations, allocatable blocks are restricted to those that match the exact size of memory requested from the allocator.
\end{figure}

\begin{figure}[!htb]
%\centering
\includegraphics[scale=.55]{figs/split.pdf}
\caption{block splitting}
Splitting of a free block creates two new blocks: a block that is the size that the allocator requested, and the other contains the remainder of a new free block with the remainder of memory.
\label{fig:split}
\end{figure}

\begin{figure}[!htb]
%\centering
\includegraphics[scale=.55]{figs/fuse.pdf}
\caption{block fusing}
An example of fusing blocks together. This creates single larger block, decreasing the number of smaller blocks unable to split.
\label{fig:fuse}
\end{figure}

%
%

### 2014-04-25 tech-project/specify_led.tex
adding led demo tex

end LED_Telos_Demo;
\end{verbatim}
%smaller file sizes
%reusable code
%only have to translate a file once
% be able to say that the code we have is the code that was verified.
% our approach takes less ROM. Its not less efficient, it is differently efficient.
Prior to using the LED component developed in the previous sections, we first must bind our \texttt{LED\_Template} specification with an appropriate realization. This is accomplished via the facility declaration located directly beneath the \texttt{uses} clause, which pairs the specification (\texttt{LED\_Template}) with a realization (\texttt{Std\_Led\_Realiz}). Note that the ability to toggle is added \textit{on top} of this facility declaration in a similar fashion. The remaining bulk of logic driving the application rests in the non terminating busy loop inside operation \texttt{Main}, where we use our enhancement-provided \texttt{Toggle} operation to successively turn each light within the strip on, then off.

@@

The \texttt{LED\_Telos\_Demo} facility above demonstrates these developments through its use of the ``\texttt{externally realized}" phrase. This signals to the RESOLVE compiler that the user is providing a non-native realization of \texttt{LED\_Template}, with the expectation that it conforms to the specifications dictated in the concept.

@@

In the context of embedded systems, these developments are especially useful as we can now write custom, low level  drivers of the sort required for most embedded applications (e.g. LED strip drivers) while still providing extensible, formally specified interfaces. It is our hope and intention that new (native) RESOLVE components will be layered on top of these low level, externally realized (yet specified) drivers -- eventually reaching a level of abstraction where we can concern ourselves exclusively with verified, native RESOLVE components.

>>>>

end LED_Telos_Demo;
\end{verbatim}
%smaller file sizes
%reusable code
%only have to translate a file once
% be able to say that the code we have is the code that was verified.
% our approach takes less ROM. Its not less efficient, it is differently efficient.
Prior to using the LED component developed in the previous sections, we first must bind our \texttt{LED\_Template} specification with an appropriate realization. This is accomplished via the facility declaration located directly beneath the \texttt{uses} clause, which pairs the specification (\texttt{LED\_Template}) with a realization (\texttt{Std\_Led\_Realiz}). Note that the enhanced ability of toggling lights is added \textit{on top} of this facility declaration in a similar fashion. The remaining bulk of logic driving the application rests in the non terminating busy loop inside operation \texttt{Main}, where we use our enhancement-provided \texttt{Toggle} operation to successively turn each light within the strip on, then off.

@@

The \texttt{LED\_Telos\_Demo} facility above demonstrates these developments through its use of the ``\texttt{externally realized}" phrase. This signals to the RESOLVE compiler that the user is providing a non-native realization of the \texttt{LED\_Template}, with the expectation that it conforms to the specifications dictated in the concept.

@@

In the terms of embedded systems, these developments are especially as it allows us to write custom low level of the sort required for most embedded applications (e.g. LED strip drivers) while still providing formally specified interfaces that might later form the foundation . It is our hope and intention that new (native) resolve components will be layered on top of these low level, externally realized (yet specified) drivers -- eventually reaching a level of abstract where we can concern ourselves exclusively with verified, native RESOLVE components.
